Extended Data

### 2021-11-16 doc/manuscript/Dow_manuscript_Nature.Rmd
rearrange methods text on climate data

@@

For the tree core analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].

<!-- Determining critical temperature windows -->

The critical temperature window (Fig. `r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
The variable and time window yielding the biggest $\Delta AICc$ are selected for our critical temperature window. Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`). We selected the critical temperature window that had the lowest $\Delta AICc$.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.

<!--Hypothesis testing -->

Correlation between the dendroband-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and of individual tree. 
We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019]. In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

@@

<!--QUILT PLOT METHODS-->
Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.

>>>>

<!-- Determining critical temperature windows -->
The critical temperature window (Fig. `r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
The variable and time window yielding the biggest $\Delta AICc$ are selected for our critical temperature window. Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`). We selected the critical temperature window that had the lowest $\Delta AICc$.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.

<!--Hypothesis testing -->
Correlation between the dendroband-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and of individual tree. 
We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019]. In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

@@

@@

<!--QUILT PLOT METHODS-->
For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on the following year's growth, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.
